---
title: "Why blocking makes a difference?"
author: "Sidhant Chaudhary"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: united
    code_folding: show
    toc: yes
    toc_float: yes
    fig_caption: yes
    number_sections: yes
    df_print: tibble
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
                      warning=F,
                      message = F,
                      fig.align='center', out.width='100%', fig.asp=1)


# R setup template
rm(list = ls())

# packages and libraries

if (!require("pacman")) install.packages("pacman")

pacman::p_unload(pacman::p_loaded(), character.only = TRUE)

pacman::p_load(tidyverse, conflicted, skimr, Hmisc, knitr, ggtext,
              
               lsmeans, emmeans, multcomp, multcompView) 


# conflicts: identical function names from different packages
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")

# ggplot theme
theme_set(theme_test() + 
          theme(
            #axis.title = element_text(size = 14),
            axis.text = element_markdown(color = "black"),
            strip.background = element_rect(fill = "skyblue"),
            strip.text = element_markdown(hjust = 0),
            legend.text = element_markdown(),
            legend.title = element_markdown()
          ))
```

# Data used
The data is downloaded from https://raw.githubusercontent.com/SchmidtPaul/DSFAIR/master/data/Clewer%26Scarisbrick2001.csv 
```{r}
dataURL <- "https://raw.githubusercontent.com/SchmidtPaul/DSFAIR/master/data/Clewer%26Scarisbrick2001.csv"
dat <- read_csv(dataURL)

dat

# Raw data visualisation
dat %>%
  ggplot(aes(x = cultivar, y = yield)) + 
  geom_boxplot(alpha = 0.1) +
  stat_summary(fun = "mean", geom = "point", shape = 18, size = 4, color = "black", 
  position = position_dodge(width = 0.75), show.legend = F) +
  geom_line(aes(group = block, color = block)) +
  geom_point(aes(color = block)) + 
  labs(caption = "Black diamonds are the arithmatic means per cultivar")
```

# Importance of blocking when data is complete
## Model without block effect
```{r}
mod_no_block <- lm(yield ~ cultivar , data = dat)
anova(mod_no_block)

mean_comparisons_no_block <- mod_no_block %>% 
  emmeans(specs = ~ cultivar) %>% # get adjusted means for cultivars
  cld(adjust="tukey", Letters=letters) # add compact letter display

mean_comparisons_no_block
```
We observe that cultivar effect is not significant according to ANOVA. Consequently, no significant difference among cultivars is observed.

## Model with block effects
```{r}
mod_block <- lm(yield ~ cultivar + block, data = dat)
anova(mod_block)

mean_comparisons_block <- mod_block %>% 
  emmeans(specs = ~ cultivar) %>% # get adjusted means for cultivars
  cld(adjust="tukey", Letters=letters) # add compact letter display

mean_comparisons_block
```
However, once blocks are considered in the model, cultivar effect is significant in ANOVA and consequently, we observe differences among cultivars. 


After visualising the comparison of these two models with and without blocks, we see that mean estimates are the same, however, confidence intervals are more precise in the model with blocking, thereby, exposing the differences among cultivars. 
```{r}
mean_comparisons_no_block %>% 
  as_tibble() %>%
  mutate(blocking = "no") %>%
  bind_rows(mean_comparisons_block %>%
              as_tibble() %>%
              mutate(blocking = "yes")) %>%
  ggplot(aes(x = cultivar, y = emmean, color = blocking)) +
  geom_pointrange(aes(ymin = lower.CL, ymax = upper.CL), position = position_dodge(width = 0.5)) + 
  geom_text(aes(y = upper.CL + 1, label = .group), position = position_dodge(width = 0.5)) 
```

# Importance of blocking with missing data 
What happens when you miss data? Blocking becomes quite important also in cases of missing data. Let's explore in the above data by dropping one value
```{r}
dat_missing <- 
  dat %>%
  mutate(yield = na_if(yield, 8))

dat_missing
```
Cultivar 4 is missing its value from 2nd block. 

## Model on missing data with no blocks 
```{r}
dat_missing %>%
  group_by(cultivar) %>%
  summarise(mean_yield = mean(yield, na.rm = T))

mod_miss_no_block <- lm(yield ~ cultivar, data = dat_missing)
anova(mod_miss_no_block)
mean_comparisons_miss_no_block <- mod_miss_no_block %>% 
  emmeans(specs = ~ cultivar) %>% # get adjusted means for cultivars
  cld(adjust="tukey", Letters=letters) # add compact letter display

mean_comparisons_miss_no_block
```
The mean estimate of C4 is 8.45. 

## Model on missing data with blocks
```{r}
mod_miss_block <- lm(yield ~ cultivar + block , data = dat_missing)
anova(mod_miss_block)
mean_comparisons_miss_block <- mod_miss_block %>% 
  emmeans(specs = ~ cultivar) %>% # get adjusted means for cultivars
  cld(adjust="tukey", Letters=letters) # add compact letter display
mean_comparisons_miss_block

```
The mean estimate of C4 is 8.23. 

```{r}
dat %>%
  group_by(cultivar) %>%
  summarise(mean_yield = mean(yield, na.rm = T))

```
The mean estimate of C4 from the model with blocks is closer to the arithmatic mean of 8.3.

The plot below also shows the narrower confidence intervals alongside a better estimate of mean in case of missing data. 
```{r}
mean_comparisons_miss_no_block %>% 
  as_tibble() %>%
  mutate(blocking = "no") %>%
  bind_rows(mean_comparisons_miss_block %>%
              as_tibble() %>%
              mutate(blocking = "yes")) %>%
  ggplot(aes(x = cultivar, y = emmean, color = blocking)) +
  geom_pointrange(aes(ymin = lower.CL, ymax = upper.CL), position = position_dodge(width = 0.5)) + 
  geom_text(aes(y = upper.CL + 1, label = .group), position = position_dodge(width = 0.5)) +  
geom_text(aes(y = emmean, label = round(emmean, 2)), position = position_dodge(width = 1))
```


# Conclusion
In conclusion, blocking is important for the following reasons: 

- More precise estimates, smaller confidence intervals
- More correct estimates in case of missing data 

# Session info
```{r}

sessionInfo()
```

